---
title: "Assignment_5"
author: "Nwe Oo Mon (Nina)"
date: "2024-09-28"
output: html_document
---
## Loading the libraries and the data
The data can be downloaded from
 [github](https://github.com/Nweoomon/Assignment5_607_092924/blob/main/airlines.csv).
 
```{r setup}
library(tidyverse)
library(knitr)

# Read the CSV file, specifying missing values
file_path <- "C:\\ProgramData\\MySQL\\MySQL Server 8.0\\Uploads\\airlines.csv"
airlines_data <- read_csv(file_path, na = c("-1"))

kable(airlines_data, caption = "Complete Airlines Dataset")
```

## From wide data form to long form
```{r long_data}
airlines_long <- airlines_data %>%
  pivot_longer(cols = c("Los Angeles", "Phoenix", "San Diego", "San Francisco", "Seattle"),
               names_to = "Destination",
               values_to = "Result")

head (airlines_long)
```

## Total Flights by Airlines
```{r total_flights}

# Summarize total flights by airline
Airline_summary <- airlines_long %>%
  group_by(Airline) %>%
  summarise(total_flights = sum(Result))

kable(Airline_summary, caption = "Total Flights by Airline")

# Summarize total flights for each destination by airline
destination_summary <- airlines_long %>%
  group_by(Destination, Airline) %>%
  summarise(total_flights = sum(Result))

kable(destination_summary, caption = "Total Flights by Destination")

# Grouped bar chart to show total flights for each destination by airline
ggplot(data = destination_summary, aes(x = Destination, y = total_flights, fill = Airline)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Total Flights to Each Destination by Airline",
       x = "Destination",
       y = "Total Number of Flights",
       fill = "Airline") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
scale_fill_manual(values = c("ALASKA" = "#0077C8", "AM WEST" = "#C8102E"))
```

The graph shows that AM WEST has significantly more flights to Phoenix compared to ALASKA and also has a higher number of flights to Los Angeles and San Diego. On the other hand, ALASKA has the most flights to Seattle, and relatively higher flights to San Francisco.

# Flight Delays
```{r flight_status_airline_destination}

airlines_long <- airlines_long %>%
  group_by(Destination, Airline) %>%
  mutate(percentage = ifelse(Status == "delayed", (Result / sum(Result)) * 100, NA))

airlines_long <- airlines_long %>%
  group_by(Destination, Airline) %>%
  mutate(label_y = ifelse(Status == "delayed", 
                          lag(cumsum(Result), default = 0) + (Result / 2), 
                          NA))

print(airlines_long)

ggplot(data = airlines_long, aes(x = Destination, y = Result, fill = Status)) +
  geom_bar(stat = "identity", width = 0.7) + 
  geom_text(data = airlines_long %>% filter(Status == "delayed"),
            aes(label = paste0(round(percentage, 1), "%"), y = label_y),
            color = "black", size = 4) +
  facet_grid(~ Airline) +
  theme_minimal() +
  labs(title = "Comparison of Delays and On-Time Flights by Airline and Destination",
       x = "Destination",
       y = "Total Flights",
       fill = "Status") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1))  
```

Alaska Airline experiences the highest percentage of delays for flights to San Francisco at 16.9%, followed by Seattle at 14.2%. Delays for Phoenix and San Diego are relatively low.
AM West Airline shows the most significant percentage of delays for flights to San Francisco (28.7%), followed by Seattle (23.3%) and Los Angeles (8.5%). Phoenix has low delay percentages.

```{r flight_status_airline}
# Summarize flight status by airline
status_summary <- airlines_long %>%
  group_by(Status, Airline) %>%
  summarise(total_flights = sum(Result))

status_summary <- status_summary %>%
  group_by(Airline) %>%
  mutate(percentage = ifelse(Status == "delayed", (total_flights / sum(total_flights)) * 100, NA))

print(status_summary)

# Create a stacked bar chart_airlines
ggplot(data = status_summary, aes(x = Airline, y = total_flights, fill = Status)) +
  geom_bar(stat = "identity", width = 0.5) +
  geom_text(data = status_summary %>% filter(Status == "delayed"),
            aes(label = paste0(round(percentage, 1), "%")),
            position = position_stack(vjust = 9.5),
            color = "black", size = 4) +
  theme_minimal() +
  labs(title = "Comparison of Delays and On-Time Flights for Airlines",
       x = "Airline",
       y = "Number of Flights",
       fill = "Flight Status") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

ALASKA Airline has 501 delays and 3274 on time flights, and delays accounted for 13.3% of their total flights.
AM WEST Airline has 787 delayed flights and 6438 on time flights, and delayed flights represented 10.9% of their total flights.

```{r flight_status_destination}
# Summarize flights for each destination by flight status
destination_status <- airlines_long %>%
  group_by(Status, Destination) %>%
  summarise(total_flights = sum(Result))

destination_status <- destination_status %>%
  group_by(Destination) %>%
  mutate(percentage = ifelse(Status == "delayed", (total_flights / sum(total_flights)) * 100, NA))

print(destination_status)

# Create a stacked bar chart_destination
ggplot(data = destination_status, aes(x = Destination, y = total_flights, fill = Status)) +
  geom_bar(stat = "identity", width = 0.5) +
  theme_minimal() +
  labs(title = "Comparison of Delays and On-Time Flights by destination",
       x = "Destination",
       y = "Number of Flights",
       fill = "Flight Status") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

San Francisco has the highest percentage of delayed flights at 21.9%, followed by Seattle at 15.2%.
Phoenix has the lowest percentage of delayed flights, with only 7.8% of flights being delayed.

## Summary

Comparing flight performance city-by-city versus overall reveals that certain cities experience more delays. AM WEST performs well overall but has a high delay rate in San Francisco and a low rate in Phoenix. 
